---
VignetteBuilder: knitr
Suggests: BiocStyle, knitr, rmarkdown
title: "Report EN1 ko transcriptome analysis, dimension reduction"
author: "Sebastian Schmidt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/sschm/OneDrive - Helmholtz Zentrum München/Manuscript Engrailed/sina_transcriptome/R_En1ko')

#load libraries

library(ggbiplot)
library(DT)
library(ggplot2)
library(Rtsne)
library(RColorBrewer)
library(pheatmap)
library(DESeq2)
library(gridExtra)
library(kableExtra)
library(biomaRt)
library(readxl)
library(devEMF)
library(devtools)
library(tidyverse)

#specifiy location of input files
length_file <- "data/featurelength_en1koexport.tabular"
count_matrix_file <- "data/countmatrix.txt"

sample_summary_file <- "data/samples.xlsx"


```

# Loading of data and normalization

***

## Count matrix

First, we load the count matrix, and compute TPM normalization.

```{r, echo = FALSE, include = FALSE, warning = FALSE}
#############################################
## Load count matrix and tidy it

# generate vector with conditions and with samples (will be used later)
sample_summary <- as.matrix(read_xlsx(sample_summary_file))
rownames(sample_summary) <- sample_summary[,1]
sample_summary <- as.data.frame(sample_summary[,-1])


# read count matrix
count_matrix <- read.table(count_matrix_file, sep = "\t", header = T, check.names = FALSE)

# assign rownames
rownames(count_matrix) <- count_matrix[,1]

# now we can get rid of first column
count_matrix <- count_matrix[,-1]

```

This is how our count matrix looks like (first 1000 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}
# we only show the table with the first 1000 rows, to speed things up.

datatable(as.data.frame(count_matrix)[1:1000,], extensions = c('Buttons'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE), caption = htmltools::tags$caption(style = 'caption-side: bottom; text-align: center;', 'Table 1: ', htmltools::em('Count matrix')))

```

This is the number of reads per sample:

```{r, echo = FALSE, include = TRUE, warning = FALSE}

#show number of reads per sample
a <- colSums(count_matrix); a

write.table(a, "results/tables/reads_per_sample.txt", sep = "\t")
```

## Transcripts per million (TPM) normalization

We compute the transcripts per million normalization. This is how it looks like (first 1000 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}

## TPM normalizations

# Read the gene - length file
feature_length <- read.table(length_file, sep = "\t", header = T, stringsAsFactors = F)

# Create a named vector with the length in kb
length_kb = feature_length$Length / 1000
names(length_kb) = feature_length$Geneid

# Divide the read counts by the length of each gene in kilobases. This gives you reads per kilobase (RPK).
matchedelements=match( rownames(count_matrix) , names(length_kb))

# we can do some checks
#length(matchedelements)
#dim(count_matrix)
#sum(is.na(matchedelements))
#count_matrix[1,]
#length_kb[58961]

# Divide matrix rows by length in kb for each gene
# Reads per kilobase (RPK) matrix
rpk_matrix <- apply(count_matrix, 2, "/", length_kb[matchedelements])

# Count up all the RPK values in a sample
rpk_factor <- colSums(rpk_matrix)

# and divide each value of the samples (columns) by this factor.
tpm_matrix <- t(apply(rpk_matrix, 1, "/", rpk_factor))

# Finally, multiply by 1 million
tpm_matrix <- tpm_matrix * 1E6




write.table(tpm_matrix, "results/tables/count_matrix_tpm.txt", sep = "\t")
```

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}
# we only show the table with the first 1000 rows, to speed things up.

datatable(as.data.frame(tpm_matrix)[1:1000,], extensions = c('Buttons'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE), caption = htmltools::tags$caption(style = 'caption-side: bottom; text-align: center;', 'Table 4: ', htmltools::em('TPM matrix')))

```

Each sample should sum up to 1-million:

```{r, echo = FALSE, include = TRUE, warning = FALSE}
colSums(tpm_matrix)
```


# Principal component analysis {.tabset}

Looks quite good. Variation between replicates of same clone is low. Separation mainly by genotype or stimulation. Only wt clone 18-3 seems to be a outlier (unstimulated dataset) and was removed for downstream analysis.

## Unstimulated samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="PCA", fig.show="hold", out.width="30%"}
#calculate pca


sample_summary_unstim <- sample_summary[,1:24]
conditions_unstim <- as.character(as.vector(sample_summary_unstim[2,]))
stimulation_unstim <- as.character(as.vector(sample_summary_unstim[1,]))
samples_unstim <- as.character(as.vector(sample_summary_unstim[5,]))

tpm_matrix_unstim <- tpm_matrix[,1:24]


tpm_matrix_unstim_pca <- tpm_matrix_unstim[rowMeans(tpm_matrix_unstim) != 0, ]
pca_unstim <- prcomp(t(na.omit(tpm_matrix_unstim_pca)), center = TRUE, scale = TRUE)

#create plots

pca_condition <- ggbiplot(pca_unstim, choices = c(1,2), ellipse=TRUE, groups=conditions_unstim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = conditions_unstim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_condition_unstimulated.pdf",pca_condition, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

pca_stimulation <- ggbiplot(pca_unstim, choices = c(1,2), ellipse=TRUE, groups=stimulation_unstim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = stimulation_unstim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_stimulation_unstimulated.pdf",pca_stimulation, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

pca_samples <- ggbiplot(pca_unstim, choices = c(1,2), ellipse=TRUE, groups=samples_unstim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = samples_unstim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_samples_unstimulated.pdf",pca_samples, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

plot(pca_condition)
plot(pca_stimulation)
plot(pca_samples)

```
```{r, include = FALSE}
#remove 18-3
count_matrix <- count_matrix[,-c(22,23,24,46,47,48)]
tpm_matrix <- tpm_matrix[,-c(22,23,24,46,47,48)]
sample_summary <- sample_summary[,-c(22,23,24,46,47,48)]


sample_summary_unstim <- sample_summary[,1:21]
conditions_unstim <- as.character(as.vector(sample_summary_unstim[2,]))
stimulation_unstim <- as.character(as.vector(sample_summary_unstim[1,]))
samples_unstim <- as.character(as.vector(sample_summary_unstim[5,]))

tpm_matrix_unstim <- tpm_matrix[,1:21]

sample_summary_stim <- sample_summary[,22:42]
conditions_stim <- as.character(as.vector(sample_summary_stim[2,]))
stimulation_stim <- as.character(as.vector(sample_summary_stim[1,]))
samples_stim <- as.character(as.vector(sample_summary_stim[5,]))

tpm_matrix_stim <- tpm_matrix[,22:42]

```


## Stimulated samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="PCA", fig.show="hold", out.width="30%"}
#calculate pca


tpm_matrix_stim_pca <- tpm_matrix_stim[rowMeans(tpm_matrix_stim) != 0, ]
pca_stim <- prcomp(t(na.omit(tpm_matrix_stim_pca)), center = TRUE, scale = TRUE)

#create plots

pca_condition <- ggbiplot(pca_stim, choices = c(1,2), ellipse=TRUE, groups=conditions_stim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = conditions_stim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_condition_stimulated.pdf",pca_condition, device = "pdf", height = 15, width = 15, units = "cm")

pca_stimulation <- ggbiplot(pca_stim, choices = c(1,2), ellipse=TRUE, groups=stimulation_stim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = stimulation_stim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_stimulation_stimulated.pdf",pca_stimulation, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

pca_samples <- ggbiplot(pca_stim, choices = c(1,2), ellipse=TRUE, groups=samples_stim, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = samples_stim)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_samples_stimulated.pdf",pca_samples, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

plot(pca_condition)
plot(pca_stimulation)
plot(pca_samples)

```

## Comparison - Stimulated vs unstimulated samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="PCA", fig.show="hold", out.width="30%"}
#calculate pca


conditions <- as.character(as.vector(sample_summary[2,]))
stimulation <- as.character(as.vector(sample_summary[1,]))
samples <- as.character(as.vector(sample_summary[5,]))

tpm_matrix_pca <- tpm_matrix[rowMeans(tpm_matrix) != 0, ]
pca <- prcomp(t(na.omit(tpm_matrix_pca)), center = TRUE, scale = TRUE)

#create plots

pca_condition <- ggbiplot(pca, choices = c(1,2), ellipse=TRUE, groups=conditions, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = conditions)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_condition_comparison.pdf",pca_condition, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

pca_stimulation <- ggbiplot(pca, choices = c(1,2), ellipse=TRUE, groups=stimulation, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = stimulation)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_stimulation_comparison.pdf",pca_stimulation, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

pca_samples <- ggbiplot(pca, choices = c(1,2), ellipse=TRUE, groups=samples, var.axes = FALSE) +
  geom_jitter(size = 2.5, aes(color = samples)) + 
  theme_classic() +
  theme(axis.text.x = element_text(color="black"),
        axis.ticks = element_line(color="black"),
        axis.text.y = element_text(color="black"))+
  xlim(-2.5,2.5)+
  ylim(-2.5,2.5)
ggsave("results/plots/PCA_samples_comparison.pdf",pca_samples, device = "pdf", height = 15, width = 15, units = "cm", dpi = 400)

plot(pca_condition)
plot(pca_stimulation)
plot(pca_samples)

```

# Clustering and correlation between sample 

Next, we perform clustering and correlation analysis, and plot these results in a couple of heatmaps.

## Unstimulated samples {.tabset}

### Distance matrix, followed by clustering

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="Heatmap of distances"}
#calculate distances between samples
sampleDists <- dist(t(tpm_matrix_unstim))

#tidy matrix
sampleDistMatrix <- as.matrix(sampleDists)

#make a vector with colors and then plot heatmap
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
annotation_col = data.frame(
  Stimulation = stimulation_unstim, 
  Condition = conditions_unstim
)
rownames(annotation_col) = colnames(tpm_matrix_unstim)

dist_heatmap_tpm <- pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors,
         annotation_col = annotation_col)

emf("results/plots/dist_heatmap_tpm_unstimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=20, height=20, family= "Arial") 
dist_heatmap_tpm
dev.off()


```

### Correlation between samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, , fig.cap="Heatmap of correlations"}

#calculate correlations between samples
sampleDists <- cor(tpm_matrix_unstim)

#tidy matrix
sampleDistMatrix <- as.matrix(sampleDists)


#make a vector with colors and then plot heatmap
colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)
cor_heatmap_tpm <- pheatmap(sampleDistMatrix, col=colors,  annotation_col = annotation_col)

emf("results/plots/cor_heatmap_tpm_unstimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=20, height=20, family= "Arial") 
cor_heatmap_tpm
dev.off()

```

### Hierarchical clustering of the samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="Hierarchical clustering"}
###################################
#simple hierarchical clustering using distances
clusters <- hclust(dist(t(tpm_matrix_unstim)))
plot(clusters, labels = colnames(tpm_matrix_unstim))

emf("results/plots/hirachical_clustering_unstimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=15, height=5, family= "Arial") 
plot(clusters, labels = colnames(tpm_matrix_unstim))
dev.off()

```
## Stimulated samples {.tabset}

### Distance matrix, followed by clustering

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="Heatmap of distances"}
#calculate distances between samples
sampleDists <- dist(t(tpm_matrix_stim))

#tidy matrix
sampleDistMatrix <- as.matrix(sampleDists)

#make a vector with colors and then plot heatmap
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
annotation_col = data.frame(
  Stimulation = stimulation_stim, 
  Condition = conditions_stim
)
rownames(annotation_col) = colnames(tpm_matrix_stim)

dist_heatmap_tpm <- pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors,
         annotation_col = annotation_col)

emf("results/plots/dist_heatmap_tpm_stimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=20, height=20, family= "Arial") 
dist_heatmap_tpm
dev.off()


```

### Correlation between samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, , fig.cap="Heatmap of correlations"}

#calculate correlations between samples
sampleDists <- cor(tpm_matrix_stim)

#tidy matrix
sampleDistMatrix <- as.matrix(sampleDists)


#make a vector with colors and then plot heatmap
colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)
cor_heatmap_tpm <- pheatmap(sampleDistMatrix, col=colors,  annotation_col = annotation_col)

emf("results/plots/cor_heatmap_tpm_stimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=20, height=20, family= "Arial") 
cor_heatmap_tpm
dev.off()

```

### Hierarchical clustering of the samples

```{r, echo = FALSE, include = TRUE, warning = FALSE, fig.cap="Hierarchical clustering"}
###################################
#simple hierarchical clustering using distances
clusters <- hclust(dist(t(tpm_matrix_stim)))
plot(clusters, labels = colnames(tpm_matrix_stim))

emf("results/plots/hirachical_clustering_stimulated.emf", emfPlus= FALSE, coordDPI = 400, 
    width=15, height=5, family= "Arial") 
plot(clusters, labels = colnames(tpm_matrix_stim))
dev.off()

```

