---
VignetteBuilder: knitr
Suggests: BiocStyle, knitr, rmarkdown
title: "Report EN1 ko transcriptome analysis, differential expression analysis for unstimulated samples"
author: "Sebastian Schmidt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/sschm/OneDrive - Helmholtz Zentrum München/Manuscript Engrailed/sina_transcriptome/R_En1ko')
#load libraries

library(DT)
library(ggplot2)
library(Rtsne)
library(RColorBrewer)
library(pheatmap)
library(DESeq2)
library(gridExtra)
library(kableExtra)
library(biomaRt)
library(readxl)
library(devEMF)
library(EnhancedVolcano)
library(xlsx)

#specifiy location of input files
length_file <- "data/featurelength_en1koexport.tabular"
count_matrix_file <- "data/countmatrix.txt"

sample_summary_file <- "data/samples.xlsx"




```

# Loading of data

***

## Count matrix

First, we load the count matrix.

```{r, echo = FALSE, include = FALSE, warning = FALSE}
#############################################
## Load count matrix and tidy it

# generate vector with conditions and with samples (will be used later)
sample_summary <- as.data.frame(read_xlsx(sample_summary_file))
rownames(sample_summary) <- sample_summary[,1]
sample_summary <- sample_summary[,-1]

sample_summary <- sample_summary[,-c(22,23,24,46,47,48)] #remove clone 18-3

# read count matrix
count_matrix <- read.table(count_matrix_file, sep = "\t", header = T, check.names = FALSE)

# assign rownames
rownames(count_matrix) <- count_matrix[,1]

# now we can get rid of first column
count_matrix <- count_matrix[,-1]

count_matrix <- count_matrix[,-c(22,23,24,46,47,48)] #remove clone 18-3


#filter for unstimulated
count_matrix <- t(count_matrix)
sample_summary <- as.data.frame(t(sample_summary))

count_matrix <- t(count_matrix[sample_summary$stimulation == "unstim",])
sample_summary <- t(sample_summary[sample_summary$stimulation == "unstim",])

```

This is how the count matrix for only unstimulated samples looks like (first 1000 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(count_matrix)[1:1000,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

This is how the sample matrix for only unstimulated samples looks like

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(sample_summary), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

# Differential expression analysis for wt vs EN1 ko samples

We performed a differential gene expression analysis using DESeq2.

This is how our experimental design table looks like. We will perform a DEG analysis treating all replicates as independent samples (design = ~condition), as well as a nested approach (design = ~clone.n + clone.n:replicate + condition) to compensate for dependencies within the groups.
```{r, echo = FALSE, include = TRUE, warning = FALSE}
#############################################
#############################################
## Differential expression analyses

sample_names <- as.character(as.vector(colnames(sample_summary)))
conditions <- as.character(as.vector(sample_summary[2,]))
stimulation <- as.character(as.vector(sample_summary[1,]))
samples <- as.character(as.vector(sample_summary[5,]))
clone <- as.character(as.vector(sample_summary[3,]))
replicate <- as.character(as.vector(sample_summary[4,]))
clone.n <- as.character(as.vector(sample_summary[6,]))
#make a data frame with the experimental design
expDesign = data.frame(sample = colnames(count_matrix), 
                       condition = conditions, 
                       clone = clone,
                       replicate = replicate,
                       clone.n = clone.n)                    
kable(expDesign) %>% kable_styling(bootstrap_options = "striped", full_width = F)
```


```{r, echo = FALSE, include = FALSE, warning = FALSE}
#Make DESEQ2 object
dds_nested <- DESeqDataSetFromMatrix(countData = count_matrix, colData = expDesign, design = ~clone.n + clone.n:replicate + condition)
#m1 <- model.matrix(~condition, colData(dds))
#mm <- model.matrix(~clone + condition, colData(dds))
#colnames(count_matrix)
#mm <- mm[,1:11]

#colnames(mm)
#unname(mm)
#all.zero <- apply(mm, 2, function(x) all(x==0)); all.zero

#idx <- which(all.zero)
#mm <- mm[,-idx]
#unname(mm)


#dim(count_matrix)
#filtering out lowly expressed genes
filter = 10
keep <- apply(count_matrix, 1, min) > filter
dds_nested <- dds_nested[keep,]




# DESEQ2 analysis
#dds <- DESeq(dds, full = mm)
dds_nested <- DESeq(dds_nested)
#get table of differentially expressed genes

#select two conditions to contrast
res_hom_nested = results(dds_nested, contrast = c("condition", "ko_hom", "ctrl"))
res_hom_nested <- res_hom_nested[order(res_hom_nested$log2FoldChange),]

```


```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}
# We want to add a column with the gene symbol and a description to table

mart = useDataset("hsapiens_gene_ensembl", useEnsembl(biomart="ensembl", version=98))


genes <- sapply(strsplit(rownames(res_hom_nested),"[.]"), `[`, 1)
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id", "hgnc_symbol", "description"),values=genes,mart=mart)
res_hom_nested$symbol = G_list[match(genes, G_list$ensembl_gene_id),]$hgnc_symbol
res_hom_nested$description = G_list[match(genes, G_list$ensembl_gene_id),]$description
write.xlsx(res_hom_nested, file = "results/tables/DEGs_full_unstim.xlsx", sheetName = "DEGs_unstim", append = TRUE)


# Filter the results absFC > 1 & P-value < 0.05

diff_exp_genes_hom_nested = res_hom_nested[which((res_hom_nested$padj < 0.05)),]

```


And these are the results of our differential gene expression analysis for homozygous EN1 ko clones vs Ctrl clones (nested approach). Only genes with a significant adjusted p-value are shown.

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}

datatable(as.data.frame(diff_exp_genes_hom_nested), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))

write.table(diff_exp_genes_hom_nested, "results/tables/DEGs_unstim.txt", sep = "\t")
write.xlsx(diff_exp_genes_hom_nested, file = "results/tables/DEGs_unstim.xlsx", sheetName = "DEGs_unstim", append = TRUE)


```


# Visualization of DEGs {.tabset .tabset-fade .tabset-pills}

Next, we employ a number of visualization methods. 

## MA-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "CairoSVG"}

plotMA(res_hom_nested, alpha = 0.01, main="DESeq2 MA plot, alpha=0.01")
```

## Vulcano-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "CairoSVG"}

#label <- c("EN1", "EN2", "FZD5", "SFRP1", "LEF1", "FZD7", "WNT7A")
label <- c("EN1", "EN2", "SP5", "DLK1", "SLC16A3", "EPHA3", "HOXB9", "OLIG2", "MAOA", "SPARCL1")

Volcano <- EnhancedVolcano(toptable = res_hom_nested,
                title = "",
                subtitle = "",
                lab = res_hom_nested$symbol, 
                x = "log2FoldChange", y = "padj", 
                
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 2.0,
                gridlines.major = FALSE,
                colConnectors = "black",
                selectLab = label,
                labCol = "black",
                drawConnectors = TRUE
) 
#ylim = c(0,15), xlim = c(-4,4),

Volcano

emf("results/plots/Volcano_unstim.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=10, family= "Arial") 
Volcano
dev.off()
```

## Heatmap

```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "CairoSVG"}
#################################

# Lets create some annotations for the columns
mat_col <- data.frame(condition = conditions)
rownames(mat_col) <- sample_names

# List with colors for each annotation.
mat_colors <- list(condition = c("orange", "darkblue"))
names(mat_colors$condition) <- c("ctrl", "ko_hom")

# TPM Matrix / Transform to log2 values

feature_length <- read.table(length_file, sep = "\t", header = T, stringsAsFactors = F)

length_kb = feature_length$Length / 1000
names(length_kb) = feature_length$Geneid

matchedelements=match( rownames(count_matrix) , names(length_kb))

rpk_matrix <- apply(count_matrix, 2, "/", length_kb[matchedelements])

rpk_factor <- colSums(rpk_matrix)

tpm_matrix <- t(apply(rpk_matrix, 1, "/", rpk_factor))

tpm_matrix <- tpm_matrix * 1E6

tpm_matrixL2 <- log2(tpm_matrix+1)

tpm_matrixL2 <- tpm_matrixL2[match(rownames(diff_exp_genes_hom_nested), rownames(tpm_matrixL2)),]

# Scale the rows
tpm_matrixL2=t(scale(t(tpm_matrixL2)))
rownames(tpm_matrixL2) = NULL

# Plot the heatmap
heatmap <- pheatmap(tpm_matrixL2, annotation_col=mat_col,  annotation_colors = mat_colors,cluster_rows = T, cluster_cols = F)


emf("results/plots/heatmap_unstim_new.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=5, family= "Arial") 
heatmap
dev.off()

```
