---
VignetteBuilder: knitr
Suggests: BiocStyle, knitr, rmarkdown
title: "Report EN1 ko transcriptome analysis, differential expression analysis unstimulated vs stimulated"
author: "Sebastian Schmidt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: TRUE
---

```{r setup, include=FALSE}
#if you want to change the look of your report, just browse to https://bootswatch.com/ and select a theme. Then type the name of the theme above, instead of 'cosmo'!
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/sschm/OneDrive - Helmholtz Zentrum München/Manuscript Engrailed/sina_transcriptome/R_En1ko')
#load libraries

library(DT)
library(ggplot2)
library(Rtsne)
library(RColorBrewer)
library(pheatmap)
library(DESeq2)
library(gridExtra)
library(kableExtra)
library(biomaRt)
library(readxl)
library(devEMF)
library(EnhancedVolcano)
library(xlsx)

#specifiy location of input files
length_file <- "data/featurelength_en1koexport.tabular"
count_matrix_file <- "data/countmatrix.txt"

sample_summary_file <- "data/samples.xlsx"

```

# Loading of data and normalization

***

## Count matrix

First, we load the count matrix.

```{r, echo = FALSE, include = FALSE, warning = FALSE}
#############################################
# Load count matrix and tidy it

# generate vector with conditions and with samples (will be used later)
sample_summary <- as.data.frame(read_xlsx(sample_summary_file))
rownames(sample_summary) <- sample_summary[,1]
sample_summary <- sample_summary[,-1]

sample_summary <- sample_summary[,-c(22,23,24,46,47,48)] #remove clone 18-3

# read count matrix
count_matrix <- read.table(count_matrix_file, sep = "\t", header = T, check.names = FALSE)

# assign rownames
rownames(count_matrix) <- count_matrix[,1]

# now we can get rid of first column
count_matrix <- count_matrix[,-1]

count_matrix <- count_matrix[,-c(22,23,24,46,47,48)] #remove clone 18-3


#filter for Ctrl and EN1 ko
count_matrix <- t(count_matrix)
sample_summary <- as.data.frame(t(sample_summary))

count_matrix_ctrl <- t(count_matrix[sample_summary$condition == "ctrl",])
sample_summary_ctrl <- t(sample_summary[sample_summary$condition == "ctrl",])

count_matrix_ko_hom <- t(count_matrix[sample_summary$condition == "ko_hom",])
sample_summary_ko_hom <- t(sample_summary[sample_summary$condition == "ko_hom",])

count_matrix <- t(count_matrix)
sample_summary <- t(sample_summary)

```

This is how the full count matrix looks like (first 100 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(count_matrix)[1:100,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

This is how the full sample matrix looks like

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(sample_summary), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```



This is how the count matrix for the Ctrl samples - unstimulated and Wnt stimulated - looks like (first 100 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(count_matrix_ctrl)[1:100,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

This is how the sample matrix for the Ctrl samples - unstimulated and Wnt stimulated - looks like

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(sample_summary_ctrl), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```



This is how the count matrix for the EN1 ko samples - unstimulated and Wnt stimulated - looks like (first 100 rows, to speed things up):

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(count_matrix_ctrl)[1:100,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

This is how the sample matrix for the EN1 ko samples - unstimulated and Wnt stimulated - looks like

```{r, echo = FALSE, include = TRUE, warning = FALSE}
datatable(as.data.frame(sample_summary_ctrl), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```


# Differential expression analysis for ctrl lines - unstimulated vs. Wnt stimulated
***
We performed a differential gene expression analysis using DESeq2.

## Differential expression analysis with DESeq2

This is how our experimental design table looks like for the Ctrl samples - unstimulated and Wnt stimulated. 
```{r, echo = FALSE, include = TRUE, warning = FALSE}
#############################################
#############################################
## Differential expression analyses

sample_names_ctrl <- as.character(as.vector(colnames(sample_summary_ctrl)))
conditions_ctrl <- as.character(as.vector(sample_summary_ctrl[2,]))
stimulation_ctrl <- as.character(as.vector(sample_summary_ctrl[1,]))
samples_ctrl <- as.character(as.vector(sample_summary_ctrl[5,]))
clone_ctrl <- as.character(as.vector(sample_summary_ctrl[3,]))
replicate_ctrl <- as.character(as.vector(sample_summary_ctrl[4,]))
clone.n_ctrl <- as.character(as.vector(sample_summary_ctrl[6,]))
#make a data frame with the experimental design
expDesign_ctrl = data.frame(sample = colnames(count_matrix_ctrl), 
                       stimulation = stimulation_ctrl, 
                       clone = clone_ctrl,
                       replicate = replicate_ctrl,
                       clone.n = clone.n_ctrl)                    
kable(expDesign_ctrl) %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

We will perform a DEG analysis using a nested approach (design = ~clone.n + clone.n:replicate + stimulation) to compensate for dependencies / replicates within the groups.

```{r, echo = FALSE, include = FALSE, warning = FALSE}
#Ctrl
dds_ctrl_nested <- DESeqDataSetFromMatrix(countData = count_matrix_ctrl, colData = expDesign_ctrl, design = ~clone.n + clone.n:replicate + stimulation)

filter = 10
keep <- apply(count_matrix_ctrl, 1, min) > filter
dds_ctrl_nested <- dds_ctrl_nested[keep,]

dds_ctrl_nested <- DESeq(dds_ctrl_nested)

res_ctrl_nested = results(dds_ctrl_nested, contrast = c("stimulation", "stim", "unstim"))
res_ctrl_nested <- res_ctrl_nested[order(res_ctrl_nested$log2FoldChange),]


```


```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}
# We want to add a column with the gene symbol and a description to table

mart = useDataset("hsapiens_gene_ensembl", useEnsembl(biomart="ensembl", version=98))

#ctrl
genes <- sapply(strsplit(rownames(res_ctrl_nested),"[.]"), `[`, 1)
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id", "hgnc_symbol", "description"),values=genes,mart=mart)
res_ctrl_nested$symbol = G_list[match(genes, G_list$ensembl_gene_id),]$hgnc_symbol
res_ctrl_nested$description = G_list[match(genes, G_list$ensembl_gene_id),]$description
#write.xlsx(res_ctrl_nested, file = "results/tables/DEGs_unstim-vs-stim_full.xlsx", sheetName = "Ctrl_unstim-vs-stim")

# Filter the results P-value < 0.05
diff_exp_genes_ctrl_nested = res_ctrl_nested[which(res_ctrl_nested$padj < 0.05),]
#write.xlsx(diff_exp_genes_ctrl_nested, file = "results/tables/DEGs_unstim-vs-stim.xlsx", sheetName = "Ctrl_unstim-vs-stim")
```


And these are the results of our differential gene expression analysis for the Ctrl samples - unstimulated and Wnt stimulated (top 100 DEGs). Only genes with a significant adjusted p-value and no fold change threshold are shown.

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}

datatable(as.data.frame(diff_exp_genes_ctrl_nested)[1:100,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))

```

## Visualization of DEGs {.tabset .tabset-fade .tabset-pills}

Next, we employ a number of visualization methods

### MA-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}

plotMA(res_ctrl_nested, alpha = 0.01, main="DESeq2 MA plot, alpha=0.01")

```

### Vulcano-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}

label <- c("EN1", "EN2", "SP5", "DLK1", "SLC16A3", "EPHA3", "HOXB9", "OLIG2", "MAOA", "SPARCL1")
Volcano <- EnhancedVolcano(toptable = res_ctrl_nested,
                title = "",
                subtitle = "",
                lab = res_ctrl_nested$symbol, 
                x = "log2FoldChange", y = "padj", 
                
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 2.0,
                gridlines.major = FALSE,
                colConnectors = "black",
                selectLab = label,
                labCol = "black",
                drawConnectors = TRUE
) 

Volcano

emf("results/plots/Volcano_unstim-vs-stim_ctrl.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=10, family= "Arial") 
Volcano
dev.off()
```

### Heatmap

```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}
# Lets create some annotations for the columns
mat_col <- data.frame(stimulation = stimulation_ctrl)
rownames(mat_col) <- sample_names_ctrl

# List with colors for each annotation.
mat_colors <- list(stimulation = c("red","darkblue"))
names(mat_colors$stimulation) <- c("unstim","stim")

# TPM Matrix / Transform to log2 values
feature_length <- read.table(length_file, sep = "\t", header = T, stringsAsFactors = F)

length_kb = feature_length$Length / 1000
names(length_kb) = feature_length$Geneid

matchedelements=match( rownames(count_matrix_ctrl) , names(length_kb))

rpk_matrix <- apply(count_matrix_ctrl, 2, "/", length_kb[matchedelements])

rpk_factor <- colSums(rpk_matrix)

tpm_matrix <- t(apply(rpk_matrix, 1, "/", rpk_factor))

tpm_matrix <- tpm_matrix * 1E6


tpm_matrixL2 <- log2(tpm_matrix+1)

tpm_matrixL2 <- tpm_matrixL2[match(rownames(diff_exp_genes_ctrl_nested), rownames(tpm_matrixL2)),]

# Scale the rows
tpm_matrixL2=t(scale(t(tpm_matrixL2)))
rownames(tpm_matrixL2) = NULL

# Plot the heatmap
heatmap <- pheatmap(tpm_matrixL2, annotation_col=mat_col,  annotation_colors = mat_colors,cluster_rows = T, cluster_cols = F)


emf("results/plots/heatmap_unstim-vs-stim_ctrl.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=5, family= "Arial") 
heatmap
dev.off()

```


# Differential expression analysis for EN1 ko lines - unstimulated vs. Wnt stimulated
***
We performed a differential gene expression analysis using DESeq2.

## Differential expression analysis with DESeq2

This is how our experimental design table looks like for the EN1 ko samples - unstimulated and Wnt stimulated.

```{r, echo = FALSE, include = TRUE, warning = FALSE}
#############################################
#############################################
## Differential expression analyses

sample_names_ko_hom <- as.character(as.vector(colnames(sample_summary_ko_hom)))
conditions_ko_hom <- as.character(as.vector(sample_summary_ko_hom[2,]))
stimulation_ko_hom <- as.character(as.vector(sample_summary_ko_hom[1,]))
samples_ko_hom <- as.character(as.vector(sample_summary_ko_hom[5,]))
clone_ko_hom <- as.character(as.vector(sample_summary_ko_hom[3,]))
replicate_ko_hom <- as.character(as.vector(sample_summary_ko_hom[4,]))
clone.n_ko_hom <- as.character(as.vector(sample_summary_ko_hom[6,]))
#make a data frame with the experimental design
expDesign_ko_hom = data.frame(sample = colnames(count_matrix_ko_hom), 
                       stimulation = stimulation_ko_hom, 
                       clone = clone_ko_hom,
                       replicate = replicate_ko_hom,
                       clone.n = clone.n_ko_hom)                    
kable(expDesign_ko_hom) %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

We will perform a DEG analysis using a nested approach (design = ~clone.n + clone.n:replicate + stimulation) to compensate for dependencies / replicates within the groups.

```{r, echo = FALSE, include = FALSE, warning = FALSE}

#ko_hom
dds_ko_hom_nested <- DESeqDataSetFromMatrix(countData = count_matrix_ko_hom, colData = expDesign_ko_hom, design = ~clone.n + clone.n:replicate + stimulation)

filter = 10
keep <- apply(count_matrix_ko_hom, 1, min) > filter
dds_ko_hom_nested <- dds_ko_hom_nested[keep,]

dds_ko_hom_nested <- DESeq(dds_ko_hom_nested)

res_hom_nested = results(dds_ko_hom_nested, contrast = c("stimulation", "stim", "unstim"))
res_hom_nested <- res_hom_nested[order(res_hom_nested$log2FoldChange),]


```


```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}
#hom
genes <- sapply(strsplit(rownames(res_hom_nested),"[.]"), `[`, 1)
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id", "hgnc_symbol", "description"),values=genes,mart=mart)
res_hom_nested$symbol = G_list[match(genes, G_list$ensembl_gene_id),]$hgnc_symbol
res_hom_nested$description = G_list[match(genes, G_list$ensembl_gene_id),]$description
#write.xlsx(res_hom_nested, file = "results/tables/DEGs_unstim-vs-stim_full.xlsx", sheetName = "EN1ko_unstim-vs-stim", append = TRUE)

# Filter the results P-value < 0.05
diff_exp_genes_hom_nested = res_hom_nested[which(res_hom_nested$padj < 0.05),]
#write.xlsx(diff_exp_genes_hom_nested, file = "results/tables/DEGs_unstim-vs-stim.xlsx", sheetName = "EN1ko_unstim-vs-stim", append = TRUE)

```

And these are the results of our differential gene expression analysis for homozygous EN1 ko clones unstimulated vs stimulated (top 100 DEGs). Only genes with a signficant adjusted p-value and no fold change threshold are shown.

```{r, echo = FALSE, include = TRUE, warning = FALSE, message=FALSE}

datatable(as.data.frame(diff_exp_genes_hom_nested)[1:100,], extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))

```


## Visualization of DEGs {.tabset .tabset-fade .tabset-pills}

Next, we employ a number of visualization methods. 

### MA-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}

plotMA(res_hom_nested, alpha = 0.01, main="DESeq2 MA plot, alpha=0.01")
```

### Vulcano-plot
```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}

Volcano <- EnhancedVolcano(toptable = res_hom_nested,
                title = "",
                subtitle = "",
                lab = res_hom_nested$symbol, 
                x = "log2FoldChange", y = "padj", 
                
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 2.0,
                gridlines.major = FALSE,
                colConnectors = "black",
                selectLab = label,
                labCol = "black",
                drawConnectors = TRUE
) 

Volcano

emf("results/plots/Volcano_unstim-vs-stim_EN1ko.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=10, family= "Arial") 
Volcano
dev.off()
```

### Heatmap

```{r, echo = FALSE, include = TRUE, warning = FALSE, dev = "jpeg"}
# Lets create some annotations for the columns
mat_col <- data.frame(stimulation = stimulation_ko_hom)
rownames(mat_col) <- sample_names_ko_hom

# List with colors for each annotation.
mat_colors <- list(stimulation = c("red","darkblue"))
names(mat_colors$stimulation) <- c("unstim","stim")

# TPM Matrix / Transform to log2 values
feature_length <- read.table(length_file, sep = "\t", header = T, stringsAsFactors = F)

length_kb = feature_length$Length / 1000
names(length_kb) = feature_length$Geneid

matchedelements=match( rownames(count_matrix_ko_hom) , names(length_kb))

rpk_matrix <- apply(count_matrix_ko_hom, 2, "/", length_kb[matchedelements])

rpk_factor <- colSums(rpk_matrix)

tpm_matrix <- t(apply(rpk_matrix, 1, "/", rpk_factor))

tpm_matrix <- tpm_matrix * 1E6

tpm_matrixL2 <- log2(tpm_matrix+1)
tpm_matrixL2 <- tpm_matrixL2[match(rownames(diff_exp_genes_hom_nested), rownames(tpm_matrixL2)),]

# Scale the rows
tpm_matrixL2=t(scale(t(tpm_matrixL2)))
rownames(tpm_matrixL2) = NULL

# Plot the heatmap
heatmap <- pheatmap(tpm_matrixL2, annotation_col=mat_col,  annotation_colors = mat_colors,cluster_rows = T, cluster_cols = F)


emf("results/plots/heatmap_unstim-vs-stim_EN1ko.emf", emfPlus= FALSE, coordDPI = 400, 
    width=7, height=5, family= "Arial") 
heatmap
dev.off()

```
