---
title: "Report EN1 ko transcriptome analysis, Pathway enrichment analysis using DEGs for Ctrl clones - unstimulated vs Wnt stimulated"
author: "Sebastian Schmidt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/sschm/OneDrive - Helmholtz Zentrum München/Manuscript Engrailed/sina_transcriptome/R_En1ko')

library(knitr)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(devEMF)
library(extrafont)
library(ggplot2)
library(org.Hs.eg.db)
library(rmarkdown)
library(DT)
library(RColorBrewer)
library(gridExtra)
library(kableExtra)
library(biomaRt)
library(ggnewscale)
library(ReactomePA)
library(xlsx)

d <- read.xlsx(file = "results/tables/DEGs_unstim-vs-stim.xlsx", sheetName = "Ctrl_unstim-vs-stim")

gene <- d$symbol
```


# Enriched GO terms 


```{r,echo = FALSE, warning=FALSE, message=FALSE}
try(ego_CC <- enrichGO(gene          = gene,
                
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                keyType = "SYMBOL",
                pAdjustMethod = "fdr",
                
                qvalueCutoff  = 0.05,
                readable      = FALSE))
try(ego_BP <- enrichGO(gene          = gene,
                  
                   OrgDb         = org.Hs.eg.db,
                   ont           = "BP",
                   keyType = "SYMBOL",
                   pAdjustMethod = "fdr",
                   
                   qvalueCutoff  = 0.05,
                   readable      = FALSE))
try(ego_MF <- enrichGO(gene          = gene,
                   
                   OrgDb         = org.Hs.eg.db,
                   ont           = "MF",
                   keyType = "SYMBOL",
                   pAdjustMethod = "fdr",
                   
                   qvalueCutoff  = 0.05,
                   readable      = FALSE))
try(ego_ALL <- enrichGO(gene          = gene,
                   
                   OrgDb         = org.Hs.eg.db,
                   ont           = "ALL",
                   keyType = "SYMBOL",
                   pAdjustMethod = "fdr",
                   
                   qvalueCutoff  = 0.05,
                   readable      = FALSE))

write.xlsx(ego_ALL, file = "results/tables/pathways_Ctrl_unstim-vs-stim.xlsx", sheetName = "GO", row.names = FALSE)

```

## Tables {.tabset}
### CC

```{r,echo = FALSE, warning=FALSE, message=FALSE}
datatable(as.data.frame(ego_CC), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```
### biological process

```{r,echo = FALSE, warning=FALSE, message=FALSE}
datatable(as.data.frame(ego_BP), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

### molecular fuction

```{r,echo = FALSE, warning=FALSE, message=FALSE}
datatable(as.data.frame(ego_MF), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))

```

### All

```{r,echo = FALSE, warning=FALSE, message=FALSE}
datatable(as.data.frame(ego_ALL), extensions = c('Buttons', 'FixedColumns'), options = list(autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1), dom = 'Bfrtip', buttons = c('csv', 'excel')))

```

##  Dotplots {.tabset} 

### CC

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=20}
dotplot(ego_CC, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### biological process 

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=20}
dotplot(ego_BP, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### molecular function 

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=20}
dotplot(ego_MF, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### ALL 

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=20}
dotplot(ego_ALL, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```


## Treeplots {.tabset} 

### CC
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
p <- treeplot(pairwise_termsim(ego_CC), showCategory = 20, cluster.params = list(n = 8)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
```


### Biological process
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=15}

p <- treeplot(pairwise_termsim(ego_BP), showCategory = 20, cluster.params = list(n = 8)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
```


### Molecular function
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=15}

p <- treeplot(pairwise_termsim(ego_MF), showCategory = 20, cluster.params = list(n = 8)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
```


### ALL
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
p <- treeplot(pairwise_termsim(ego_ALL), showCategory = 20, cluster.params = list(n = 8)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
emf("results/plots/Ctrl_unstim-vs-stim_eGO_ALL.emf",width = 10, height = 8, coordDPI = 400, family = "Arial") 
p
dev.off()
```


## Emapplot {.tabset} 
### CC
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ego_CC), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### BP
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ego_BP), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### MF
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ego_MF), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

### ALL
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ego_ALL), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```




# Enriched KEGG terms

Gene identifier changed from HGNC symbol to EntrezID
                
```{r,echo = FALSE, warning=FALSE, message=FALSE}
hs <- org.Hs.eg.db
test <- select(hs, 
       keys = gene,
       columns = c("ENTREZID"),
       keytype = "SYMBOL")
geneEntrez <- test$ENTREZID
ekegg <- enrichKEGG(gene         = geneEntrez,
                   pvalueCutoff = 0.05)

write.xlsx(ekegg, file = "results/tables/pathways_Ctrl_unstim-vs-stim.xlsx", sheetName = "KEGG", row.names = FALSE, append = TRUE)
```

Here are the enriched KEGG terms

## Table
```{r,echo = FALSE}
datatable(as.data.frame(ekegg), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

## Dotplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=5}
dotplot(ekegg, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

## Treeplot

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=15}

p <- treeplot(pairwise_termsim(ekegg), showCategory = 20, cluster.params = list(n = 10)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
emf("results/plots/Ctrl_unstim-vs-stim_eKEGG.emf",width = 10, height = 8, coordDPI = 400, family = "Arial") 
p
dev.off()
```

## Emapplot

```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ekegg), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

# Enriched Wikipathway terms

Gene identifier changed from HGNC symbol to EntrezID

```{r,echo = FALSE, warning=FALSE, message=FALSE }
ewp <- enrichWP(geneEntrez, organism = "Homo sapiens")

write.xlsx(ewp, file = "results/tables/pathways_Ctrl_unstim-vs-stim.xlsx", sheetName = "WP", row.names = FALSE, append = TRUE)
```
Here are the enriched WP terms

## Table
```{r,echo = FALSE}
datatable(as.data.frame(ewp), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

## Dotplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=5}
dotplot(ewp, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

## Treeplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=5}

p <- treeplot(pairwise_termsim(ewp), showCategory = 20, cluster.params = list(n = 5)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
emf("results/plots/Ctrl_unstim-vs-stim_eWP.emf",width = 10, height = 8, coordDPI = 400, family = "Arial") 
p
dev.off()

```

## Emapplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ewp), showCategory = 200) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

# Enriched Reactome terms

Gene identifier changed from HGNC symbol to EntrezID

```{r,echo = FALSE, warning=FALSE, message=FALSE}

ereactome <- enrichPathway(gene=geneEntrez, pvalueCutoff = 0.05, readable=TRUE)


#write.xlsx(ereactome, file = "results/tables/pathways_Ctrl_unstim-vs-stim.xlsx", sheetName = "Reactome", row.names = FALSE, append = TRUE)
```

Here are the enriched Reactome terms

## Table
```{r,echo = FALSE}
datatable(as.data.frame(ereactome), extensions = c('Buttons', 'FixedColumns'), options = list(dom = 'Bfrtip', buttons = c('csv', 'excel'), autoWidth = TRUE, scrollX = TRUE, fixedColumns = list(rightColumns = 1)))
```

## Dotplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}
dotplot(ereactome, showCategory = 20) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```

## Treeplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=10}

p <- treeplot(pairwise_termsim(ereactome), showCategory = 20, cluster.params = list(n = 8)) + scale_color_gradientn(name = "p.adjust", colours = c( "red", "blue"))

p
emf("results/plots/Ctrl_unstim-vs-stim_eReactome.emf",width = 10, height = 8, coordDPI = 400, family = "Arial") 
p
dev.off()
```

## Emapplot
```{r,echo = FALSE, warning=FALSE, message=FALSE, fig.width=30, fig.height=30}
emapplot(pairwise_termsim(ereactome), showCategory = 150) + scale_fill_gradientn(name = "p.adjust", colours = c( "red", "blue"))
```


